---
title: "Introduction to thyroidBRS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to thyroidBRS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(thyroidBRS)
```

## Background

Papillary thyroid carcinoma (PTC) is predominantly driven by either a
`BRAF`^V600E^ mutation or a RAS-family mutation. Agrawal et al. (2014)
(*Cell*, [doi:10.1016/j.cell.2014.09.050](https://doi.org/10.1016/j.cell.2014.09.050))
showed that these two driver groups have distinguishable transcriptional
profiles, and defined the **BRAF-RAS Score (BRS)**: a continuous score from
-1 (most BRAF-like) to +1 (most RAS-like), computed from a 71-gene expression
signature (their Figure S7A).

The original study only reported a BRS for the 391 (of 496) tumors that had
**both** exome and RNA sequencing data — exome data was needed to define
which tumors belonged to the BRAF-mutant reference group and which to the
RAS-mutant reference group, but **not** to score a tumor once those two
reference groups (and their expression centroids) are fixed. Any additional
sample with RNA-seq expression data can be scored against those same
centroids, with no mutation calling required.

`thyroidBRS` implements exactly that two-step idea:

1. **`brs_fit()`** — given a cohort with known BRAF-mutant/RAS-mutant labels
   for at least some samples, computes the two reference centroids.
2. **`predict()`** — scores *any* sample's expression against those
   centroids, including samples that were never labeled.

## A worked example

`thyroidBRS` ships with the 71-gene signature ([brs_genes]) but not any real
expression data (the underlying TCGA data is large and best fetched with
`TCGAbiolinks` directly). This vignette instead uses a small simulated
cohort so the example runs quickly and self-contained; the same calls work
unchanged on real data.

```{r simulate-data}
set.seed(1)
genes <- na.omit(brs_genes$current_symbol)
samples <- paste0("sample_", 1:40)

# In real data this would be log2(TPM + 1) expression, genes x samples.
expr <- matrix(
  rnorm(length(genes) * length(samples)),
  nrow = length(genes), dimnames = list(genes, samples)
)

# Simulate 24 samples with a known BRAF-mutant/RAS-mutant label...
labels <- setNames(rep(c("Braf-like", "Ras-like"), each = 12), samples[1:24])

# ...and 16 with expression only (as if their exome data were unavailable).
unlabeled_samples <- samples[25:40]
```

### Fit the reference centroids

```{r fit}
fit <- brs_fit(expr, labels)
fit
```

### Score the labeled cohort, as a sanity check

```{r validate}
preds_labeled <- predict(fit, expr[, names(labels)])
validate_brs(preds_labeled, labels)
```

With this simulated (random, unstructured) data, concordance will be close
to chance — with a real, biologically-driven cohort like TCGA-THCA, the
concordance against the published labels is 98.7% (386/391 samples; see
`?brs_fit` and the package README for details on how that was measured).

### Score the previously-unlabeled samples

This is the actual point of the package: extending classification beyond
the original exome-sequenced cohort.

```{r classify-new}
preds_new <- predict(fit, expr[, unlabeled_samples])
head(preds_new)
```

### One-step alternative

`brs_score()` wraps `brs_fit()` + `predict()` for the common case of scoring
the same cohort you fit on:

```{r one-shot}
all_preds <- brs_score(expr, labels)
head(all_preds)
```

## Using real TCGA data

To reproduce the BRS on TCGA-THCA:

```r
library(TCGAbiolinks)
# Download STAR - Counts RNA-seq for TCGA-THCA (Primary Tumor), build a
# log2(TPM + 1) matrix keyed by gene symbol, one column per patient — see
# your own preprocessing pipeline for the exact steps.
#
# Then pull the published BRAF_RAF_class as your reference labels:
subtype_table <- TCGAquery_subtype(tumor = "thca")
labels <- setNames(as.character(subtype_table$BRAF_RAF_class), subtype_table$patient)
labels[trimws(labels) == ""] <- NA  # "" means unclassified in this table, not a real label

fit <- brs_fit(expr, labels)
preds <- predict(fit, expr)  # scores every patient, including previously-unclassified ones
```

## Reference

Agrawal N, Akbani R, Aksoy BA, et al. Integrated Genomic Characterization
of Papillary Thyroid Carcinoma. *Cell*. 2014;159(3):676-690.
[doi:10.1016/j.cell.2014.09.050](https://doi.org/10.1016/j.cell.2014.09.050)
