---
title: "Introduction to thyroidBRS"
author: "Camila Barrios"
package: thyroidBRS
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Introduction to thyroidBRS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-knitr, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(thyroidBRS)
```

## Background

Papillary thyroid carcinoma (PTC) is predominantly driven by either a
`BRAF`^V600E^ mutation or a RAS-family mutation. Agrawal et al. (2014)
(*Cell*,
[doi:10.1016/j.cell.2014.09.050](https://doi.org/10.1016/j.cell.2014.09.050))
showed that the two driver groups have distinguishable transcriptional
profiles, and defined the **BRAF-RAS Score (BRS)** from a 71-gene signature
(their Figure S7A).

The study reported a BRS for the 391 tumors that had **both** exome and RNA
sequencing — exome data was needed to say which tumors belonged to the
BRAF-mutant reference group and which to the RAS-mutant one, but **not** to
score a tumor once those groups and their centroids are fixed. Any additional
sample with expression data can be scored against the same centroids.

`thyroidBRS` implements exactly that two-step idea:

1. **`brs_fit()`** — given a cohort with known driver mutations for at least
   some samples, computes the two reference centroids.
2. **`predict()`** — scores *any* sample against those centroids, including
   samples that were never labeled.

## How the score is defined

Extended Experimental Procedures 14.1 defines the score of a tumor `t` as the
difference between its distance to the BRAF centroid `c(B)` and to the RAS
centroid `c(R)`:

$$\mathrm{BRS}(t) = \lVert v(t) - c(B) \rVert_2 - \lVert v(t) - c(R) \rVert_2$$

where `v(t)` is the sample's expression across the signature, z-scored
gene-by-gene against the reference cohort, and the norm is normalized by the
number of genes. Negative scores are BRAF^V600E^-like, positive RAS-like.

`predict()` returns two flavours of the same quantity:

- `brs_score` — the value above. Stable per sample: it does not change when
  you score a different set of samples against the same centroids.
- `brs_scaled` — the same value rescaled so the scored set spans
  $[-1, +1]$, which is the axis the published figures use. This is a property
  of the *set*, so it moves if the cohort changes and is undefined for a
  single sample.

## What this reproduces

This package is a re-implementation from the paper's description; the original
code was never released. It reproduces the published *classification*, not the
published numbers:

| Compared against | Result |
| --- | --- |
| Driver-mutation status, 10-fold CV | 99.7% |
| Published class labels, samples held out of the fit | 96.4% |
| Published continuous score, Spearman | 0.941 |

Expression pipelines have moved since 2014 (RSEM on hg19 then, STAR on
GENCODE v36 now) and that difference is not reconstructable, so the scores
this returns are close to but not the same as the ones in the 2014
supplementary table. Classify and rank with them; do not expect them to match
value for value.

## Which samples define the reference groups

This is the easiest thing to get wrong, so it is worth being explicit.

In the paper, set B is the **BRAF-V600E-mutant** tumors and set R is the
**RAS-mutant** tumors. The groups are defined by genotype.

They are *not* the BRAF-like/RAS-like classes the BRS produces. In the
published TCGA-THCA tables those are the column `BRAF_RAF_class`, which is
exactly `sign(BRAF_RAF_score)` — the classifier's own output. Fitting
centroids on it is circular, and it pulls tumors with other drivers into both
reference groups: 43 other-driver tumors and 1 RAS-mutant tumor land in the
"BRAF" group, and 68 other-driver tumors in the "RAS" group.

On TCGA-THCA the two fits happen to end up close — the axes correlate at
0.998 and only 2 of 505 tumors change class — so the practical damage is
small there. The problem is what it does to the *evaluation*: if you fit on
the published classes and then score the same samples, agreeing with them
measures nothing.

`brs_fit()` accepts either vocabulary, so nothing breaks if you have
`"Braf-like"`/`"Ras-like"` labels, but prefer the mutation columns.

## A worked example

`thyroidBRS` ships the signature but no expression data. This example uses a
small simulated cohort so it runs quickly; the same calls work unchanged on
real data.

```{r simulate-data}
set.seed(1)
genes <- na.omit(brs_genes$current_symbol)
samples <- paste0("sample_", 1:60)

# Two groups with a real difference across the signature, so the example
# shows the classifier doing something rather than fitting noise.
group <- rep(c("BRAF_V600E", "RAS", "unknown"), c(24, 12, 24))
shift <- ifelse(brs_genes$up_in[!is.na(brs_genes$current_symbol)] == "RAS",
                1, -1)

expr <- matrix(rnorm(length(genes) * length(samples)),
               nrow = length(genes), dimnames = list(genes, samples))
expr <- expr + outer(shift, ifelse(group == "BRAF_V600E", -0.5, 0.5))

# Mutation labels for the 36 samples that have exome data...
labels <- setNames(group[1:36], samples[1:36])
# ...and 24 with expression only.
unlabeled <- samples[37:60]
```

### Fit the reference centroids

```{r fit}
fit <- brs_fit(expr, labels)
fit
```

### Score the previously-unlabeled samples

This is the point of the package: extending classification beyond the
exome-sequenced subset.

```{r classify-new}
preds_new <- predict(fit, expr[, unlabeled])
head(preds_new)
table(preds_new$brs_class)
```

### Check against the labels you do have

```{r validate}
preds_all <- predict(fit, expr)
validate_brs(preds_all, labels, fit = fit)
```

Note the `n_resubstituted` field and the warning. Every one of those 36
samples also defined the centroids, so this number is a resubstitution
estimate, not a validation. It is a useful smoke test — if it is *not* high,
something is wrong — but it is not evidence that the classifier generalizes.
How optimistic is it? At the dimensions of the real TCGA analysis (70 genes,
391 samples split 272/119), resubstitution on data with no signal whatsoever
still lands around 68%.

For a real estimate, hold samples out:

```{r crossval}
set.seed(2)
fold <- sample(rep(1:4, length.out = length(labels)))
cv <- character(length(labels))
names(cv) <- names(labels)

for (k in 1:4) {
  f <- brs_fit(expr, labels[fold != k])
  held <- names(labels)[fold == k]
  cv[held] <- predict(f, expr[, held, drop = FALSE])$brs_class
}

truth <- ifelse(labels == "BRAF_V600E", "Braf-like", "Ras-like")
mean(cv == truth)
```

### One-step alternative

`brs_score()` wraps `brs_fit()` + `predict()` for the common case of scoring
the same cohort you fit on:

```{r one-shot}
all_preds <- brs_score(expr, labels)
head(all_preds)
```

## Expression containers

`brs_fit()`, `predict()` and `brs_score()` also take a
`SummarizedExperiment` or an `ExpressionSet`, so data from `TCGAbiolinks`
or `GEOquery` goes in without unwrapping:

```r
se <- GDCprepare(q)                      # several assays
fit <- brs_fit(se, "paper_BRAF.RAS", assay = "tpm_unstranded",
               log2_transform = TRUE)
```

Two things to watch. Pick the assay explicitly — the first assay of a GDC
object is raw counts, which is not what you want, and `brs_fit()` warns if
you leave the choice to it. And `labels` may be given as a single string, in
which case it is read as a column of `colData()` / `pData()`.

## Microarray data, and what does not transfer

The signature behaves well on microarrays. On GSE33630 (Affymetrix HG-U133
Plus 2.0, 49 PTCs) 68 of the 70 genes are present, the first principal
component of the signature accounts for 45% of the variance, and 67 of those
68 genes load on it with the sign their Figure S7A block predicts. The
biology is intact.

What does not survive a platform change is the **zero threshold**. A
`brs_fit` stores the per-gene means and SDs of the cohort it was fitted on;
feed it data from a different platform and those recentre the samples
somewhere else entirely. Scoring GSE33630 against TCGA RNA-seq centroids
ranks the tumors essentially perfectly — Spearman 0.98 against the array's
own signature axis — and then calls all 49 of them BRAF-like.

So:

- **Same platform and pipeline as the fit** — everything works, and the
  frozen standardization is what makes the score robust. Scoring a cohort
  that is 100% RAS-like still gets 97.5% of them right, because nothing about
  the score depends on the composition of the set being scored.
- **A different platform** — fit new centroids on that platform if you have
  driver-mutation labels for it. If you do not, use `brs_score` as a
  *ranking* and do not trust the class call. `predict()` warns when every
  sample lands on the same side of zero, which is the usual symptom.

Re-standardizing against the new cohort instead is tempting and does fix the
cross-platform threshold, but it buys that by making every sample's score
depend on the company it keeps: applied to a cohort that happens to be all
RAS-like, it gets 8% of them right. That is why it is not the default.

## Using real TCGA data

`data-raw/validate_tcga.R` in the package sources runs the full reproduction.
The shape of it:

```r
library(TCGAbiolinks)

# 1. A log2(TPM + 1) matrix of primary tumors, gene symbols x patient barcode.
q <- GDCquery(project = "TCGA-THCA",
              data.category = "Transcriptome Profiling",
              data.type = "Gene Expression Quantification",
              workflow.type = "STAR - Counts")
GDCdownload(q)
se <- GDCprepare(q)
# keep assay "tpm_unstranded", primary tumors (-01), collapse duplicated
# gene symbols, then log2(x + 1)  ->  expr

# 2. Reference labels: driver mutation, NOT the published class.
ref <- read.csv(system.file("extdata", "thca_reference.csv",
                            package = "thyroidBRS"))
mut <- setNames(ref$driver_group, ref$patient)
mut <- mut[mut %in% c("BRAF_V600E", "RAS")]

# 3. Fit, then score everyone.
fit <- brs_fit(expr, mut)
preds <- predict(fit, expr)

# 4. Validate against something that did not enter the fit.
published <- setNames(ref$published_class, ref$patient)
validate_brs(preds, published[!is.na(published)], fit = fit)
```

Doing this today scores 505 primary tumors — 114 more than the
published 391 — and agrees with the published score at Spearman 0.941,
with 96.4% class concordance on the 111 tumors that were not part of the
reference groups.

## Session info

```{r session-info}
sessionInfo()
```

## Reference

Agrawal N, Akbani R, Aksoy BA, et al. Integrated Genomic Characterization
of Papillary Thyroid Carcinoma. *Cell*. 2014;159(3):676-690.
[doi:10.1016/j.cell.2014.09.050](https://doi.org/10.1016/j.cell.2014.09.050)
